### 2023-12-15 submission/manuscript.tex
Editing / submitted

The ability to generate millions of 16S rRNA gene amplicon and
metagenomic sequence reads has allowed researchers to multiplex multiple
samples on the same sequencing run by pooling separate PCRs that can be
deconvoluted later based on index (aka barcode) sequences that are
embedded into the sequence of the PCR primers (1--3). Unfortunately, it
is common to observe variation in the number of sequence reads from each
sample vary by as much as 100-fold (e.g, see Figure S1). This occurs
because pooling of DNA from multiple PCRs is fraught with numerous
opportunities for technical errors to compound leading to a skewed
distribution. Aside from developing better methods of pooling DNA, the
question of how to control for uneven sequencing effort in microbial
ecology studies has become controversial.

@@

In 2014, McMurdie and Holmes (11) announced that ``rarefying'' of
microbial community data was ``statistically inadmissible'' because it
omits valid data. In their simulations, they observed that rarefying
reduced the statistical power to correctly cluster samples into the same
treatment groups based on beta diversity metrics. Although the detail
was lost on many subsequent researchers, they did not describe
rarefaction, but rarefying (12). According to McMurdie and Holmes,
rarefying is a single subsampling of an OTU abundance table whereas
rarefaction repeats the subsampling step many times. Furthermore, their
simulations penalized the rarefied data by removing 15\% of the samples
and evaluated clustering quality using a clustering algorithm that
performed worse for rarefied data. According to my reanalysis of
rarefaction using the original simulation framework, rarefaction
outperformed other normalization method being used in 2014 for both
alpha and beta diversity metrics (12). Others have also critiqued the
original work (13). Regardless, there is a general sense in the
community that rarefying and rarefaction should not be used with
microbiome data.

Since McMurdie and Holmes published their simulations, others have
developed alternative approaches to control for uneven sequencing effort
in amplicon sequencing studies. For alpha diversity metrics, Willis used
toy datasets to demonstrate that one could estimate the richness for
each sample in a dataset and use those values for statistical
comparisons (14). Non-parametric estimators of richness and diversity
(15, 16) and parametric estimators of richness (17) have been used in
microbial ecology studies. For beta diversity metrics at least four
approaches have been pursued. First, one could use relative abundance
values where the observed number of sequences in an OTU is divided by
the total number of sequences in the sample (18). Second, normalization
strategies have been developed where the relative abundance is
multiplied by the size of the minimum desired sequencing effort and
fractional values are reapportioned to the OTUs to obtain integer values
(19, 20). Third, a variety of center log-ratio methods have been
developed where the compositional nature of the OTU counts is removed
and used to calculated Euclidean distances (aka Aitchison distances)
(18, 20--23). This strategy is purported to control for uneven
sequencing effort (22, 24); however, some have noticed that this feature
breaks down under certain conditions (25). Finally, as mentioned above,
variance stabilization transformations have been recommended to generate
values that can be used to calculate Euclidean distances (11).

The ongoing controversy over the use of rarefaction and the recent
development of alternative strategies to control for uneven sequencing
effort caused me to question how these methods compared to each other
using a simulation framework that overcame the issues with the McMurdie
and Holmes study. My analysis included 16S rRNA gene sequence data from
from 12 studies that characterized the variation in bacterial
communities from diverse environments (Table 1 and Figure S1). The
sequences were assigned to OTUs using a standard pipeline and their
frequencies and the number of sequences found in each sample were used
to generate simulated communities and treatment effects. For each
dataset and simulation, 100 replicate datasets were generated and used
as input to each of the strategies for controlling for uneven sequencing
effort. My overall conclusion was that rarefaction outperformed the
alternative strategies.

@@

\textbf{\emph{Without rarefaction, metrics of alpha diversity are
sensitive to sequencing effort.}} To test the sensitivity of various
approaches of measuring alpha diversity to sequencing effort, I
generated null models using OTU counts and sequencing depths from 12
studies. Under a null model, each community from the same dataset would
be expected to have the same alpha diversity regardless of the
sequencing effort. I measured the richness of the communities in each
dataset without any correction, using scaled ranked subsampling (SRS)
normalized OTU counts, with estimates based on non-parametric and
parametric approaches, and using rarefaction (e.g.~Figure S2). For each
dataset, all of the approaches, except for rarefaction, showed a strong
correlation between richness and the number of sequences in the sample
(Figure 1A). Next, I assessed diversity using the Shannon diversity
index and the inverse Simpson diversity index without any correction,
using normalized OTU counts, and rarefaction; I also used a
non-parametric estimator of Shannon diversity. The correlation between
sequencing depth and the diversity metric was not as strong as it was
for richness and the inverse Simpson diversity values were less
sensitive than the Shannon diversity values; however, the correlation to
the diversity metrics calculated with rarefaction were the lowest for
all of the metrics and studies (Figure 1A). The alpha-diversity metrics
calculated with rarefaction consistently demonstrated a lack of
sensitivity to sequencing depth.

\textbf{\emph{Without rarefaction, metrics of beta diversity are
sensitive to sequencing effort.}} To test the sensitivity of various
approaches of measuring beta diversity to sequencing effort, I used the
same null models used for studying the sensitivity of alpha diversity.
Under a null model, the ecological distance between any pair of samples
would be the same regardless of the difference in the number of
sequences observed in each sample (e.g., Figure S3). First, I calculated
the Jaccard distance coefficient between all pairs of communities within
a dataset. The Jaccard distance coefficient is the fraction of OTUs that
are unique to either community and does not account for the abundance of
the OTUs. Jaccard distances were calculated using the uncorrected OTU
counts, with rarefaction, relative abundances, and following
normalization using cumulative sum scaling (CSS) and SRS. Only the
distances calculated with rarefaction showed a lack of sensitivity to
sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the
Bray-Curtis distance coefficient, which is a popular metric that
incorporates the abundance of each OTU. Similar to what I observed with
the Jaccard coefficient, only the data normalized with rarefaction
showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I
calculated the Euclidean distance on raw OTU counts where the central
log-ratio (CLR) was calculated (i.e., Aitchison distances) using matrix
completion to replace zero counts (Robust CLR), adding a pseudo-count of
one to all OTU counts prior to calculating the CLR (One CLR), adding a
pseudo-count of one divided by the total number of sequences obtained
for the community (Nudge CLR), and imputing the value of zero counts
(Zero CLR). The Aitchison distances were all strongly sensitive to
sequencing effort (Figure 1B). Finally, I used the
variance-stabilization transformation (VST) from DESeq2 prior to
calculating Euclidean distances. Again, there was a strong sensitivity
to sequencing effort (Figure 1B). Although Euclidean distances are not
typically used on raw or rarefaction-normalized count data in ecology
(26), Euclidean distances calculated with rarefaction were not sensitive
to sequencing effort. Across each of the beta diversity metrics and
approaches used to account for uneven sequencing effort and sparsity,
rarefaction was the least sensitive approach to differences in
sequencing effort.

\textbf{\emph{Rarefaction limits the detection of false positives when
sequencing effort and treatment group are confounded.}} Next, I
investigated the impact of the various strategies and metrics on falsely
detecting a significant difference using the the same communities
generated from the null model in the analysis of alpha and beta
diversity metrics. To test for differences in alpha and beta diversity I
used the non-parametric Wilcoxon test and non-parametric
permutation-based multivariate analysis of variance (PERMANOVA). First,
I employed an unbiased null treatment model to measure the false
detection rate, which should not have meaningfully differed from 5\%.
Indeed, for each dataset and alpha and beta diversity metric and
strategy for accounting for uneven sequencing, approximately 5\% of the
tests yielded a significant result (Figure 2). Second, I employed a
biased null treatment model where the treatment group was completely
confounded with the number of sequences in each sample (Figure S4).
Under this model, only the data normalized with rarefaction consistently
resulted in a 5\% false positive rate for alpha and beta diversity
metrics. I then used a less severe bias where the smallest 5\% of
samples were assigned to one treatment group and the largest 5\% were
assigned to another; the other samples were randomly assigned to both
treatment groups. Although the sensitivity of the false positive rate to
the level of confounding varied across methods of quantifying alpha and
beta diversity, rarefaction was the only method to consistently control
the false positive rate (Figure 3). These results aligned with the
observed sensitivity of alpha and beta diversity metrics to sequencing
effort and underscore the value of rarefaction.

\textbf{\emph{Rarefaction preserves the statistical power to detect
differences between treatment groups.}} To assess the impact of
different approaches to control for uneven sequencing effort I performed
two additional simulations. In the first simulation, I implemented a
skewed abundance distribution model to create two treatment groups for
each dataset that were each populated with half of the samples each with
the same number of sequences as the samples had in the observed data.
The two treatment groups varied in their structure such that one had the
same abundances as the null distribution above and the other had 10\% of
its OTUs randomly selected to increase their counts by 5\%. The power to
detect differences in richness between the two simulated treatment
groups by all approaches was low (Figure 4A). This was likely because
the approach for generating the perturbed community did not necessarily
change the number of OTUs in each treatment group. Regardless, the
simulations testing differences in richness using the Rice and Stream
datasets had the greatest power when the richness data were calculated
with rarefaction. To explore this further, a richness-adjusted community
model was created by removing 3\% of the OTUs from a null model model.
As suggested by the first simulation, the richness data calculated with
rarefaction had a higher statistical power than the other approaches
when measuring richness (Figure 5). The simulations testing the power to
detect differences in Shannon diversity also showed that rarefaction
performed better than the other methods (Figure 4A). When testing for
differences in the Inverse Simpson diversity index the the difference
between rarefaction and the other methods was negligible (Figure 4A).
For tests of beta diversity I found that rarefaction was the most
reliable approach to maintain statistical power to detect differences
between two communities. Among the tests using the Jaccard and
Bray-Curtis metrics, raw count data and CSS normalized data had little
power relative to using rarefaction, relative abundance, and SRS to
normalize the uneven sequencing depths. The differences in power between
counts normalized with rarefaction, relative abundance, and SRS data was
small, but if there were differences, the power obtained using
rarefaction was greater than the other methods. Among the tests using
Euclidean distances, using raw counts and CLR and DESeq2 transformed
data had little power relative to the distances calculated using
rarefaction and relative abundances. This power-based analysis of the
simulated communities using different methods of handling uneven sample
sizes demonstrated the value of rarefaction for preserving the
statistical power to detect differences between treatment groups for
measures of alpha and beta diversity.

\textbf{\emph{Increased rarefaction depth reduces intra-sample variation
in alpha and beta diversity.}} Once concern with using rarefaction is
the perceived loss of sequencing information when a large fraction of
data appears to be removed when the community with the greatest
sequencing depth is sampled to the size of the community with the least
(e.g., the smallest sample in the Bioethanol dataset had 1.04\% of
sequences that were in the largest sample). To assess the sensitivity of
alpha and beta diversity metrics to rarefaction depth, I again used the
dataset generated using the null models, but used rarefaction with each
community to varying sequencing depths (Figure 6). The richness values
increased with sequencing effort as rare OTUs would continue be
detected. In contrast, the Shannon diversity and Bray-Curtis values
plateaued with increased sequencing effort. This result was expected
since increased sequencing would lead to increased precision in the
measured relative abundance of the OTUs. Next, I measured the
coefficient of variation (i.e., the mean divided by the standard
deviation) between samples for richness, Shannon diversity, and
Bray-Curtis distances. Although the mean richness appeared to increased
unbounded with sequencing effort, the coefficient of variation for each
dataset was relatively stable. In general, the coefficient of variation
increased slightly with sequencing depth only to decline once smaller
samples were removed from the analysis at higher sequencing depths.
Interestingly, the coefficient of variation between Shannon diversity
values decreased towards zero with increased sequencing effort and the
coefficient of variation between Bray-Curtis distances tended to
increase slightly for samples where the standard deviation did not
decrease as quickly as the mean distance between samples. Regardless,
the coefficients of variation were relatively small. This analysis
indicates that there are benefits to increased sequencing depths.

\textbf{\emph{Most studies have a high level of sequencing coverage.}}
To explore the concern over loss of sequencing depth further, I
calculated the Good's coverage for the observed data. The median
coverage for each dataset ranged between 89.4 and 99.8\% for the
Seagrass and Human datasets, respectively (Figure 7). When I used a
rarefaction threshold with each dataset at the size of the smallest
community in the dataset, with the exception of the Seagrass, Rice, and
Stream datasets, the median coverage with rarefaction was still greater
than 90\%. These results suggest that most studies had a level of
sequencing coverage that aligned with the diversity of the communities.
Next, I used the null model for each dataset to ask how much sequencing
effort was required to obtain higher levels of coverage. To obtain 95
and 99\% coverage, an average of 2.70 and 101.20-fold more sequence data
was estimated to be required than was required to obtain 90\% coverage,
respectively (Figure 7). As suggested by the simulated coverages curve
in Figure 7, these estimates are conservative. Regardless, the
sequencing effort required to achieve higher sequencing depth would
likely limit the number of samples that could be sequenced when
controlling for costs. Although it may be disconcerting to use
rarefaction to normalize to a sequencing depth that is considerably
lower than that obtained for the best sequenced community in a dataset,
sequencing coverage for many environments is probably adequate even at
the lower sequencing depth. Of course, regardless of the concerns
surrounding the choice of the rarefaction depth, the results throughout
this study demonstrate that rarefaction is necessary to avoid reaching
incorrect inferences.

@@

Over the past decade, the question of whether to use rarefaction with
microbial community sequence data has become controversial. The analyses
I presented here strongly indicate that rarefaction is necessary to
control for uneven sequencing effort when comparing communities based on
alpha and beta diversity indices. Compared to all other methods,
rarefaction removed the correlation between sequencing depth and alpha
or beta diversity metrics when the sequencing depth varied by as much as
97-fold across samples. I showed that this correlation could lead
falsely detecting differences between treatment groups if sequencing
depth and sequencing effort are confounded. The correlation with
sequencing effort leads to an artificial increase in the variation
between samples and a reduced power to detect true differences in alpha
and beta diversity. For these reasons, rarefaction is a valuable tool to
control for uneven sequencing effort until improved statistical
procedures are developed or it becomes possible to more evenly
distribute sequencing effort across samples.

The primary alternative to rarefaction for measuring alpha diversity is
to estimate the metric using non-parametric or parametric models with
raw counts and to then compare the estimates (7, 14). My results
demonstrate that such approaches are limited for several reasons. First,
non-parametric richness estimates such as ACE and Chao1 are sensitive to
sequencing effort. Therefore, these strategies do not, in practice,
remove the effects of sequencing effort. Second, parametric approaches,
such as those implemented in the \texttt{breakaway} R package, generate
confidence intervals that are likely to include the true richness and
theoretically shrink with increased sequencing effort. Yet for most
samples, the confidence intervals around the estimates are too wide to
be meaningful, again leading to an inability to remove the effects of
sequencing effort. Third, it has become an increasingly common practice
for researchers to remove sequences that are rare in a sample (e.g.,
those that appear once). Although that approach was not taken in this
study, removing rare sequences would skew the distribution of sequences
and OTUs leading to a distortion of the measurement of alpha diversity
(27). The effects of removing rare sequences would vary across samples
depending on the number of sequences in each sample. One interesting
result of this analysis was the demonstration that as metrics that
depend less on rare taxa are used, the effect of uneven sequencing
effort was reduced. For example, richness counts a sequence appearing
once as much as sequence appearing 1,000 times, while the Shannon
diversity index places more emphasis on the more abundant sequence, and
the inverse Simpson index even more. Although normalizing communities to
a common number of sequences is also suggested (e.g., SRS normalization)
to control for uneven sequencing effort, the current analysis indicates
that its performance does not meet that of rarefaction. For analysis of
alpha diversity metrics, rarefaction is the most effective and
consistent approach to control for uneven sequencing effort.

Use of relative abundances, normalized counts, variance stabilizing
transformations, and centered log ratios have each been recommended as
more robust alternatives to rarefaction. Again, the only approach that
consistently removed the effects of uneven sequencing effort was
rarefaction. Most of the recommendations borrow techniques from methods
for identifying differential gene expression. Unfortunately, there is an
important but under appreciated difference between gene expression and
community data. This is the interpretation of unobserved data. For gene
expression analysis in a single organism the lack of any sequencing data
for a gene would indicate that although the gene was present, its
expression was below the limit of detection. Sequencing the same
organism under multiple conditions would not lead to a seemingly
unbounded number of genes in the organisms. Rather, the number of genes
has a definite limit that is knowable from the genome sequence. With
microbiome data, an unobserved sequence could mean that the organism was
present, but below the limit of detection or that the organism was
missing. Because we have yet to exhaustively sample any community in the
same way we have sequenced a single genome, it is unreliable to impute
the presence of all organisms. Yet, this is exactly what the variance
stabilization transformation and most CLR techniques do. This analysis
has demonstrated a clear correlation between distances calculated by
these methods and sequencing effort. This result is at odds with the
claims by others that the distances are scale invariant (22, 24). Again,
rarefaction is the most effective and consistent approach to control for
uneven sequencing effort when calculating beta diversity metrics.

>>>>

The ability to generate millions of 16S rRNA gene amplicon and
metagenomic sequence reads has allowed researchers to multiplex multiple
samples on the same sequencing run by pooling separate PCRs that can be
deconvoluted later based on index (aka barcode) sequences that are
embedded into the sequence of the PCR primers (1--3). Unfortunately, it
is common to observe variation in the number of sequence reads from each
sample vary by as much as 100-fold (e.g, see Figure S1). This occurs
because pooling of DNA from multiple PCRs is fraught with numerous
opportunities for random sampling errors to compound leading to a skewed
distribution. Aside from developing better methods of pooling DNA, the
question of how to control for uneven sequencing effort in microbial
ecology studies has become controversial.

@@

In 2014, McMurdie and Holmes (11) announced that ``rarefying'' of
microbial community data was ``statistically inadmissible'' because it
omits valid data. In their simulations, they observed that rarefying
reduced the statistical power to correctly cluster samples into the same
treatment groups based on beta diversity metrics. I have recently shown
that their analysis had a number of important problems (12). First,
although the detail was lost on many subsequent researchers, they did
not describe rarefaction. Instead they described rarefying as a single
subsampling of an OTU abundance table whereas rarefaction repeats the
subsampling step many times. Furthermore, their simulations penalized
the rarefied data by removing 15\% of the samples but did not remove
those samples when considernig their other methods. They also evaluated
the accuracy of clustering samples using a clustering algorithm that
performed worse for rarefied data. According to my reanalysis of
rarefaction using their original simulation framework, but using all of
the samples, rarefaction outperformed the other normalization methods
they described for both alpha and beta diversity metrics (12). Others
have also critiqued the original work (13).

Since McMurdie and Holmes published their simulations, others have
developed alternative approaches to control for uneven sequencing effort
in amplicon sequencing studies. For alpha diversity metrics, Willis used
overly simplistic datasets to demonstrate that one could estimate the
richness for each sample in a dataset and use those values for
statistical comparisons (14). Non-parametric estimators of the minimum
richness in a community and diversity (15--17) and parametric estimators
of richness (18) have been used in microbial ecology studies. One
non-parametric approach, iNEXT, combines rarefaction on samples more
than a threshold number of sequences and extrapoliation on samples below
the threshold (17). This method has become popular in traditional
ecology studies, but has not been widely used amongst microbial
ecologists. For beta diversity metrics at least four approaches have
been pursued as alternatives to rarefaction. First, one could use
relative abundance values where the observed number of sequences in an
OTU is divided by the total number of sequences in the sample (19).
Second, normalization strategies have been developed where the relative
abundance is multiplied by the size of the minimum desired sequencing
effort and fractional values are reapportioned to the OTUs to obtain
integer values (20, 21). Third, a variety of center log-ratio methods
have been developed where the compositional nature of the OTU counts is
removed and used to calculated Euclidean distances (aka Aitchison
distances) (19, 21--24). This strategy is purported to control for
uneven sequencing effort (23, 25); however, some have noticed that this
feature breaks down under certain conditions (26). Finally variance
stabilization transformations have been recommended to generate values
that can be used to calculate Euclidean distances (11).

The ongoing controversy over the use of rarefaction and the recent
development of alternative strategies to control for uneven sequencing
effort caused me to question how these methods compared to each other
using a simulation framework that overcame the issues with the McMurdie
and Holmes study (11). My analysis included 16S rRNA gene sequence data
from 12 diverse environments (Table 1 and Figure S1). The sequences were
assigned to OTUs using a standard pipeline and their frequencies and the
number of sequences found in each sample were used to generate simulated
communities and treatment effects. For each dataset and simulation, 100
replicate datasets were generated and used as input to each of the
strategies for controlling for uneven sequencing effort. My overall
conclusion was that rarefaction outperformed the alternative strategies.

@@

\textbf{\emph{Without rarefaction, metrics of alpha diversity are
sensitive to sequencing effort.}} To test the sensitivity of various
approaches of measuring alpha diversity to sequencing effort, I
generated null models using OTU counts and sequencing depths from 12
studies. Under a null model, each community from the same dataset would
be expected to have the same alpha diversity regardless of the
sequencing effort. I measured the richness of the communities in each
dataset without any correction, using scaled ranked subsampling (SRS)
normalized OTU counts, with estimates based on non-parametric and
parametric approaches, and using rarefaction (e.g.~Figure S2). For each
dataset, all of the approaches, except for rarefaction and possibly the
size-based iNEXT estimates, showed a strong correlation between richness
and the number of sequences in the sample (Figure 1A). The size-based
iNEXT estimates were correlated with sampling effort for the Marine and
Soil datasts, which had fewer samples and more uniform sequencing depth
distributions than the other datasets. Next, I assessed diversity using
the Shannon diversity index and the inverse Simpson diversity index
without any correction, using normalized OTU counts, non-parametric
estimation, and rarefaction. The correlation between sequencing depth
and the Shannon diversity metric was not as strong as it was for
richness. The inverse Simpson diversity values was not sensitive to
sampling effort (Figure 1A). The alpha-diversity metrics calculated with
rarefaction consistently demonstrated a lack of sensitivity to
sequencing depth.

\textbf{\emph{Without rarefaction, metrics of beta diversity are
sensitive to sequencing effort.}} To test the sensitivity of various
approaches of measuring beta diversity to sequencing effort, I used the
same null models used for studying the sensitivity of alpha diversity.
Under a null model, the ecological distance between any pair of samples
would be the same regardless of the difference in the number of
sequences observed in each sample (e.g., Figure S3). First, I calculated
the Jaccard distance coefficient between all pairs of communities within
a dataset. The Jaccard distance coefficient is the fraction of OTUs that
are unique to either community and does not account for the abundance of
the OTUs. Jaccard distances were calculated using the uncorrected OTU
counts, with rarefaction, relative abundances, and following
normalization using cumulative sum scaling (CSS) and SRS. Only the
distances calculated with rarefaction showed a lack of sensitivity to
sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the
Bray-Curtis distance coefficient, which is a popular metric that
incorporates the abundance of each OTU. Similar to what I observed with
the Jaccard coefficient, only the data normalized with rarefaction
showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I
calculated the Euclidean distance on raw OTU counts where the central
log-ratio (CLR) was calculated (i.e., Aitchison distances) by adding a
pseudo-count of one to all OTU counts prior to calculating the CLR (One
CLR), adding a pseudo-count of one divided by the total number of
sequences obtained for the community (Nudge CLR), imputing the value of
zero counts (Zero CLR), and using CLR on non-zero counts followed by
matrix completion on zero counts (Robust Principle Components Analysis
{[}PCA{]}). The Aitchison distances were all strongly sensitive to
sequencing effort (Figure 1B). Finally, I used the
variance-stabilization transformation (VST) from DESeq2 prior to
calculating Euclidean distances. Again, there was a strong sensitivity
to sequencing effort (Figure 1B). Although use of Euclidean distances is
not advised with raw or rarefaction-normalized count data in ecology
(27), Euclidean distances calculated with rarefaction were not sensitive
to sequencing effort. Across each of the beta diversity metrics and
approaches used to account for uneven sequencing effort and sparsity,
rarefaction was the least sensitive approach to differences in
sequencing effort.

\textbf{\emph{Rarefaction limits the detection of false positives when
sequencing effort and treatment group are confounded.}} Next, I
investigated the impact of the various strategies and metrics on falsely
detecting a significant difference using the the same communities
generated from the null model in the analysis of alpha and beta
diversity metrics. To test for differences in alpha and beta diversity I
used the non-parametric Wilcoxon test and non-parametric
permutation-based multivariate analysis of variance (PERMANOVA). First,
I employed an unbiased null treatment model to measure the false
detection rate, which should not have meaningfully differed from 5\%.
Indeed, for each dataset and alpha and beta diversity metric and
strategy for accounting for uneven sequencing, approximately 5\% of the
tests yielded a significant result (Figure 2). Second, I employed a
biased null treatment model where the treatment group was confounded
with the number of sequences in each sample such that the 50\% of
samples with the least sequencing effort were in one treatment group and
the others were in another treatment group (Figure S4). Under this
model, only the data normalized with rarefaction consistently resulted
in a 5\% false positive rate for alpha and beta diversity metrics. I
then used a less severe bias where the smallest 5\% of samples were
assigned to one treatment group and the largest 5\% were assigned to
another; the other samples were randomly assigned to both treatment
groups. Although the sensitivity of the false positive rate to the level
of confounding varied across methods of quantifying alpha and beta
diversity, rarefaction and the iNEXT size-based estimates for alpha
diveristy were the only methods to consistently control the false
positive rate (Figure 3). These results aligned with the observed
sensitivity of alpha and beta diversity metrics to sequencing effort and
underscore the value of rarefaction.

\textbf{\emph{Rarefaction preserves the statistical power to detect
differences between treatment groups.}} To assess the impact of
different approaches to control for uneven sequencing effort I performed
two additional simulations. In the first simulation, I implemented a
skewed abundance distribution model to create two treatment groups for
each dataset that were each populated with half of the samples each with
the same number of sequences as the samples had in the observed data.
The two treatment groups varied in their structure such that one had the
same abundances as the null distribution above and the other had 10\% of
its OTUs randomly selected to increase their counts by 5\%. The power to
detect differences in richness between the two simulated treatment
groups by all approaches was low (Figure 4A). This was likely because
the approach for generating the perturbed community did not necessarily
change the number of OTUs in each treatment group; the differences
primarily affected the distribution of the sequences across OTUs.
Regardless, the simulations testing differences in richness using the
Rice and Stream datasets had the greatest power when the richness data
were calculated with rarefaction. To explore this further, a
richness-adjusted community model was created by randomly selecting and
removing 3\% of the OTUs from a null model. As suggested by the first
simulation, the richness data calculated with rarefaction consistently
had the highest statistical power among the approaches used to measure
richness followed by the iNEXT size-based estimates (Figure 5).
Returning to simulations where treatment groups varied in their
structure, the power to detect differences in Shannon diversity was the
lowest when using raw relative abundances than when using the other
methods (Figures 4A and 5). When testing for differences in the Inverse
Simpson diversity index the the difference between rarefaction and the
other methods was negligible (Figures 4A and 5). For tests of beta
diversity, I found that rarefaction was the most reliable approach to
maintain statistical power to detect differences between two communities
(Figure 4B). Among the tests using the Jaccard and Bray-Curtis metrics,
raw count data and CSS normalized data had little power relative to
using rarefaction, relative abundance, and SRS to normalize the uneven
sequencing depths. The differences in power between counts normalized
with rarefaction, relative abundance, and SRS data was small, but if
there were differences, the power obtained using rarefaction was greater
than the other methods. Among the tests using Euclidean distances, using
raw counts and CLR and DESeq2 transformed data had little power relative
to the distances calculated using rarefaction and relative abundances.
This power-based analysis of the simulated communities using different
methods of handling uneven sample sizes demonstrated the value of
rarefaction for preserving the statistical power to detect differences
between treatment groups for measures of alpha and beta diversity.

\textbf{\emph{Increased rarefaction depth reduces intra-sample variation
in alpha and beta diversity.}} One concern with using rarefaction is the
perceived loss of sequencing information when a large fraction of data
appears to be removed when the community with the greatest sequencing
depth is sampled to the size of the community with the least (e.g., the
smallest sample in the Bioethanol dataset had 1.04\% of sequences that
were in the largest sample). To assess the sensitivity of alpha and beta
diversity metrics to rarefaction depth, I again used the dataset
generated using the null models, but used rarefaction with each
community to varying sequencing depths (Figure 6). The richness values
increased with sequencing effort as rare OTUs would continue be
detected. In contrast, the Shannon diversity and Bray-Curtis values
plateaued with increased sequencing effort. This result was expected
since increased sequencing would lead to increased precision in the
measured relative abundance of the OTUs. Next, I measured the
coefficient of variation (i.e., the mean divided by the standard
deviation) between samples for richness, Shannon diversity, and
Bray-Curtis distances. Although the mean richness appeared to increase
unbounded with sequencing effort, the coefficient of variation for each
dataset was relatively stable. In general, the coefficient of variation
increased slightly with sequencing depth only to decline once smaller
samples were removed from the analysis at higher sequencing depths.
Interestingly, the coefficient of variation between Shannon diversity
values decreased towards zero with increased sequencing effort and the
coefficient of variation between Bray-Curtis distances tended to
increase slightly for samples where the standard deviation did not
decrease as quickly as the mean distance between samples. Regardless,
the coefficients of variation were relatively small. This analysis
indicates that there are benefits to increased sequencing depths.

\textbf{\emph{Most studies have a high level of sequencing coverage.}}
To explore the concern over loss of sequencing depth further, I
calculated the Good's coverage for the observed data. The median
coverage for each dataset ranged between 89.4 and 99.8\% for the
Seagrass and Human datasets, respectively (Figure 7). When I used a
rarefaction threshold with each dataset at the size of the smallest
community in the dataset, with the exception of the Seagrass, Rice, and
Stream datasets, the median coverage with rarefaction was still greater
than 90\% (see the box plots in Figure 7). These results suggest that
most studies had a level of sequencing coverage that aligned with the
diversity of the communities. Next, I used the null model for each
dataset to ask how much sequencing effort was required to obtain higher
levels of coverage. To obtain 95 and 99\% coverage, an average of 3 and
101-fold more sequence data was estimated to be required than was
required to obtain 90\% coverage, respectively (Figure 7). As suggested
by the simulated coverages curve in Figure 7, these estimates are
conservative. Regardless, the sequencing effort required to achieve
higher sequencing depth would likely limit the number of samples that
could be sequenced when controlling for costs. Although it may be
disconcerting to use rarefaction to normalize to a sequencing depth that
is considerably lower than that obtained for the best sequenced
community in a dataset, sequencing coverage for many environments is
probably adequate even at the lower sequencing depth. Of course,
regardless of the concerns surrounding the choice of the rarefaction
depth, the results throughout this study demonstrate that rarefaction is
necessary to avoid reaching incorrect inferences.

@@

Over the past decade, the question of whether to use rarefaction with
microbial community sequence data has become controversial. The analyses
I presented here strongly indicate that rarefaction is necessary to
control for uneven sequencing effort when comparing communities based on
alpha and beta diversity indices. Compared to all other methods,
rarefaction was the most consistent method for removing the correlation
between sequencing depth and alpha or beta diversity metrics even when
the sequencing depth varied by as much as 97-fold across samples. I
showed that this correlation could lead falsely detecting differences
between treatment groups if sequencing depth and sequencing effort are
confounded. The correlation with sequencing effort leads to an
artificial increase in the variation between samples and a reduced power
to detect true differences in alpha and beta diversity. For these
reasons, rarefaction is a valuable tool to control for uneven sequencing
effort until improved statistical procedures are developed or it becomes
possible to more evenly distribute sequencing effort across samples.

Historically, the primary alternative to rarefaction for measuring alpha
diversity is to estimate the metric using non-parametric or parametric
models with raw counts and to then compare the estimates (7, 14). My
results demonstrate that such approaches are limited for several
reasons. First, non-parametric richness estimates such as ACE and Chao1
are sensitive to sequencing effort. Therefore, these estimators do not,
in practice, remove the effects of sequencing effort. Second, parametric
approaches, such as those implemented in the \texttt{breakaway} R
package, generate confidence intervals that are likely to include the
true richness and theoretically shrink with increased sequencing effort.
Yet for most samples, the confidence intervals around the estimates are
too wide to be meaningful, again leading to an inability to remove the
effects of sequencing effort. Third, it has become an increasingly
common practice for researchers to remove sequences that are rare in a
sample (e.g., those that appear once). Although that approach was not
taken in this study, removing rare sequences would skew the distribution
of sequences and OTUs leading to a distortion of the measurement of
alpha diversity (28). The effects of removing rare sequences would vary
across samples depending on the number of sequences in each sample. One
interesting result of this analysis was the demonstration that as
metrics that depend less on rare taxa are used, the effect of uneven
sequencing effort was reduced. For example, richness counts a sequence
appearing once as much as sequence appearing 1,000 times, while the
Shannon diversity index places more emphasis on the more abundant
sequence, and the inverse Simpson index even more. Although normalizing
communities to a common number of sequences is also suggested (e.g., SRS
normalization) to control for uneven sequencing effort, the current
analysis indicates that its performance does not meet that of
rarefaction. Because alpha diversity metrics are so sensitive to
sampling effort and residual PCR and sequencing artifacts, it is
important to treat their values on a relative (e.g., 5\% difference)
rather than absolute basis (e.g., difference of 48 OTUs). Rarefaction is
an effective and consistent approach to control for uneven sequencing
effort.

Use of relative abundances, normalized counts, variance stabilizing
transformations, and centered log ratios have each been recommended as
more robust alternatives to rarefaction. Again, the only approach that
consistently removed the effects of uneven sequencing effort across
alpha and beta diversity analyses was rarefaction. Many of the
alternative methods have been borrowed from differential gene expression
analysis. Unfortunately, there is an important but under appreciated
difference between gene expression and community data. This is the
interpretation of unobserved data. For gene expression analysis in a
single organism the lack of any sequencing data for a gene would
indicate that although the gene was present, its expression was below
the limit of detection. Sequencing the same organism under multiple
conditions would not lead to a seemingly unbounded number of genes in
the organisms. Rather, the number of genes has a definite limit that is
knowable from the genome sequence. With microbiome data, an unobserved
sequence could mean that the organism was present, but below the limit
of detection or that the organism was missing. Because we have yet to
exhaustively sample any community in the same way we have sequenced a
single genome, it is unreliable to impute the presence of all organisms.
Yet, this is exactly what the variance stabilization transformation and
most CLR techniques do. This analysis has demonstrated a clear
correlation between distances calculated by these methods and sequencing
effort. This result is at odds with the claims by others that the
distances are scale invariant (23, 25). Again, rarefaction is the most
effective and consistent approach to control for uneven sequencing
effort when calculating beta diversity metrics.

Two common critiques of rarefaction are that the approach omits valid
data and that the selection of rarefaction threshold is arbitrary (11).
I disagree with both critiques. All of the data are used to calculate
the mean value of the metrics after repeatedly subsampling the data.
When the dataset is subsampled, every sequence has a random chance of
being included in the calculated metric. When that subsampling is
repeated a large number of times (e.g., 100 or 1000) the risks of
ignoring or oversampling rare taxa are mitigated. As for the second
criticism, I would resist the claim that the selection of the
rarefaction threshold is arbitrary. In practice, there is a tradeoff
between sampling breadth and sequencing depth. Greater breadth will
increase the statistical power to compare treatment groups and greater
depth will increase the resolution to describe the communities. My
personal process for selecting a rarefaction threshold involves looking
for a natural break in the distribution of the number of sequences. For
example, the Lake dataset used in this study had a clear break around
10,000 sequences (Figure S1). I would also consider what samples are
below any threshold that I select. If there were critical samples below
the threshold I would either reduce the threshold or obtain more
sequences from those samples. As shown in my analysis of Good's coverage
values, most studies obtain an ample level of coverage and would need to
increase their sequencing depth by 10-fold to increase the coverage by
several percentage points. In past work, I have favored increasing
sampling breadth over sequencing depth.
